---
title: "Drawing Inference"
author: "Lucy D'Agostino McGowan"
footer:  "[Dr. Lucy D'Agostino McGowan](https://lucymcgowan.com)"
logo: "img/favicon.png"
editor: source
format: 
  revealjs: 
    theme: [custom.scss]
    slide-number: true
    chalkboard: true
knitr:
  opts_chunk: 
    code-fold: true
    echo: true
---

```{r, child = "setup.rmd"}
```

## Data

```{r}
#| echo: false
load("full_magnolia_data.rds")
magnolia_data <- full_magnolia_data %>%
  filter(id == 3) %>%
  select(observation, leaf_length, leaf_width)
```

```{r}
#| code-fold: false
magnolia_data 
```

## Magnolia data {.small}

```{r}
ggplot(magnolia_data, aes(x = leaf_width, y = leaf_length)) +
  geom_point() + 
  labs(x = "Leaf length (cm)",
       y = "Leaf width (cm)")
```
## Full Magnolia data {.small}

```{r}
#| code-fold: false
full_magnolia_data <- full_magnolia_data %>%
  group_by(id) %>%
  summarise(max_length = max(leaf_length),
            min_length = min(leaf_length),
            mean_length = mean(leaf_length),
            mean_width = mean(leaf_width)) %>%
  mutate(inches = ifelse(max_length < 10, 1, 0),
         inches = ifelse(min_length < 2, 1, 0),
         flipped = ifelse(mean_length < mean_width, 1, 0)) %>%
  left_join(full_magnolia_data, by = "id") %>%
  select(-max_length, -mean_length, - mean_width) %>%
  mutate(leaf_length2 = ifelse(flipped, leaf_width, leaf_length),
         leaf_width = ifelse(flipped, leaf_length, leaf_width),
         leaf_length = leaf_length2)
```

## Magnolia data {.small}

```{r}
ggplot(full_magnolia_data, aes(x = leaf_width, y = leaf_length)) +
  geom_point() + 
  geom_point(data = magnolia_data, color = "cornflower blue") + 
  labs(x = "Leaf length (cm)",
       y = "Leaf width (cm)")
```

## Magnolia data {.small}

::: {.question .small}
What if I want to know the relationship between leaf length and leaf width of the magnolias **on the Mag Quad**?
:::

. . .

<br>

::: {.question .small}
How can we quantify how much we'd expect the slope to differ from one random sample to another?
:::

> * We need a measure of **uncertainty**
> * How about the **standard error** of the slope?
> * The **standard error** is how much we expect $\hat{\beta}_1$ to vary from one random sample to another.

## Magnolia data {.small}

::: {.question .small}
How can we quantify how much we'd expect the slope to differ from one random sample to another?
:::

```{r}
#| code-fold: false
mod <- lm(leaf_length ~ leaf_width, data = magnolia_data)
summary(mod)
```

## Magnolia data {.small}

We need a **test statistic** that incorporates $\hat{\beta}_1$ and the standard error


```{r}
#| code-fold: false
mod <- lm(leaf_length ~ leaf_width, data = magnolia_data)
summary(mod)
```


> * $t = \frac{\hat{\beta}_1}{SE_{\hat{\beta}_1}}$



## Magnolia data {.small}

::: {.question .small}
How do we interpret this?
:::

```{r}
#| code-fold: false
mod <- lm(leaf_length ~ leaf_width, data = magnolia_data)
summary(mod)
```

> * "$\hat{\beta}_1$ is more than `r summary(mod) %>% .$coefficients %>% .[2,3] %>% round(2)` standard errors above a slope of zero"


## Magnolia data {.small}

::: {.question .small}
How do we know what values of this statistic are worth paying attention to?
:::

```{r}
#| code-fold: false
mod <- lm(leaf_length ~ leaf_width, data = magnolia_data)
summary(mod)
```


> * confidence intervals
> * p-values
> * Hypothesis testing: $H_0: \beta_1 = 0$ $H_A: \beta_1 \neq 0$

## Magnolia data {.small}

::: {.question .small}
How do get a confidence interval for $\hat{\beta}_1$? What function can we use in R?
:::

. . .

<br>

```{r}
#| code-fold: false
confint(mod)
```

. . .

::: {.question .small}
How do we interpret this value?
:::

## {{< fa laptop >}} `Application Exercise` {.small}

1. Open `appex-08.qmd`
2. Fit the model of `leaf_length` and `leaf_width` in your data
3. Calculate a confidence interval for the estimate $\hat\beta_1$
4. Interpret this value

```{r}
#| echo: false
countdown::countdown(5)
```

## Hypothesis testing 

> * So far, we have estimated the relationship between the length of magnolia leaves and the width.
> * This could be useful if we wanted to understand, on average, how these variables are related (**estimation**)
> * This could also be useful if we wanted to guess how long a leaf was based on how wide it is (**prediction**)
> * What if we just want to know whether there *is some* relationship bewteen the two? (**hypothesis testing**)

## Hypothesis testing

> * **Null hypothesis**: There is no relationship between leaf length and leaf width
>     * $H_0: \beta_1 = 0$
> * **Alternative hypothesis**: There **is** a relationship between leaf length and leaf width
>     * $H_A: \beta_1 \neq 0$

## Hypothesis testing

```{r}
#| code-fold: false
summary(mod)
```

. . .

::: {.question}
Is $\hat\beta_1$ different from 0?
:::

## Hypothesis testing

```{r}
#| code-fold: false
summary(mod)
```

::: {.question}
Is $\beta_1$ different from 0? (notice the lack of the hat!)
:::
  
## p-value

The probability of observing a statistic as extreme or more extreme than the observed test statistic **given the null hypothesis is true**

## p-value

```{r}
#| code-fold: false
summary(mod)
```


::: {.question}
What is the p-value? What is the interpretation?
:::

## Hypothesis testing {.small}

> * **Null hypothesis**: $\beta_1 = 0$ (there is no relationship between the width and length of a magnolia leaf)
> * **Alternative hypothesis**:  $\beta_1 \neq 0$ (there is a relationship between the width and length of a magnolia leaf)
> * Often we have an $\alpha$ level cutoff to compare the **p-value** to, for example **0.05**.
> * If **p-value** < 0.05, we **reject the null hypothesis**
> * If **p-value** > 0.05, we **fail to reject the null hypothesis**
> * Why don't we **ever** "accept" the null hypothesis?
> * **absense of evidence is not evidence of absense**

## p-value

```{r}
#| code-fold: false
summary(mod)
```

::: {.question}
Do we reject the null hypothesis?
:::

## {{< fa laptop >}} `Application Exercise` {.small}

1. Open `appex-08.qmd`
2. Examine the summary of the model of `leaf_length` and `leaf_width` with your data
3. Test the null hypothesis that there is no relationship between the length and width of magnolia leaves
4. What is the p-value? What is the result of your hypothesis test? 

```{r}
#| echo: false
countdown::countdown(2)
```

