---
title: "Types of variables"
author: "Lucy D'Agostino McGowan"
footer:  "[Dr. Lucy D'Agostino McGowan](https://lucymcgowan.com)"
logo: "img/favicon.png"
editor: source
format: 
  revealjs: 
    theme: [custom.scss]
    slide-number: true
    chalkboard: true
knitr:
  opts_chunk: 
    code-fold: true
    echo: true
---


## Variable types

```{r child = "setup.Rmd"}
```

```{r}
#| label: packages
#| echo: false
library(Stat2Data)
```


* There are two major classes of variables
  * numeric (quantitative) 
  * categorical


## Variable types

* Recall from the first week of class, you can use the `glimpse()` function to see all of your variables and their types

```{r}
#| code-fold: false
data("PorschePrice")
glimpse(PorschePrice)
```

> * What are the variables here?
> * `fct`: "factor" this is a type of categorical variable



## Variable types

* Recall from the first week of class, you can use the `glimpse()` function to see all of your variables and their types


```{r}
#| code-fold: false
#| eval: false
glimpse(starwars)
```


```{r}
#| echo: false
glimpse(starwars[, 1:5])
```


> * `chr`: "character" this is a type of categorical variable



## Variable types

> * So far, our models have only included _numeric_ (_quantitative_) variables
> * What would the equation be for predicting $y$ from $x$ when $x$ is numeric?
> * What would happen if $x$ is categorical?
>    * What would the equation be for predicting $y$ from $x$ if $x$ is categorical with 2 levels?
>    * What would the equation be for predicting $y$ from $x$ if $x$ is categorical with 3 levels?
  

## indicator variable {.center}

An **indicator variable** uses two values, usually 0 and 1, to indicate whether a data case does (1) or does not (0) belong to a specific category


## Indicator variable 

```{r}
#| echo: false
data("Diamonds")
```

::: smaller

```{r}
#| echo: false
Diamonds %>% 
  select(TotalPrice, Color, Carat) %>%
  DT::datatable()
```
:::



## Indicator variables

::: question 
What does this line of code do?
:::

```{r}
#| code-fold: false
#| code-line-numbers: "3"
Diamonds <- Diamonds %>%
  mutate(
    ColorD = ifelse(Color == "D", 1, 0), 
    ColorE = ifelse(Color == "E", 1, 0),
    ColorF = ifelse(Color == "F", 1, 0),
    ColorG = ifelse(Color == "G", 1, 0),
    ColorH = ifelse(Color == "H", 1, 0),
    ColorI = ifelse(Color == "I", 1, 0),
    ColorJ = ifelse(Color == "J", 1, 0)
  )
```




## Indicator variables

::: question
What does this line of code do?
:::

```{r}
#| code-fold: false
#| code-line-numbers: "4"
Diamonds <- Diamonds %>%
  mutate(
    ColorD = ifelse(Color == "D", 1, 0), 
    ColorE = ifelse(Color == "E", 1, 0), 
    ColorF = ifelse(Color == "F", 1, 0),
    ColorG = ifelse(Color == "G", 1, 0),
    ColorH = ifelse(Color == "H", 1, 0),
    ColorI = ifelse(Color == "I", 1, 0),
    ColorJ = ifelse(Color == "J", 1, 0)
  )
```



## Indicator variables

::: smaller

```{r}
#| echo: false
Diamonds %>%
  select(TotalPrice, Carat, Color, ColorD, ColorE, ColorF, ColorG, ColorH, ColorI) %>%
  DT::datatable()
```

:::

## Indicator variables

::: question
What if I wanted to model the relationship between `TotalPrice` and `Color`?
:::

::: smaller

```{r}
#| echo: false
Diamonds %>%
  select(TotalPrice, Carat, Color, ColorD, ColorE, ColorF, ColorG, ColorH, ColorI) %>%
  DT::datatable(options = list(pageLength = 5))
```
:::



## Indicator variables {.small}

::: question
Why is `ColorJ` `NA`?
:::


```{r}
#| code-fold: false
lm(TotalPrice ~ ColorD + ColorE + ColorF + ColorG + ColorH + ColorI + ColorJ,
   data = Diamonds)
```


> * When including indicator variables in a model for `k` categories, always include `k-1` 
> * The one that is left out is the "reference" category



## Indicator variables {.small}

::: question
What is the reference category?
:::


```{r}
#| code-fold: false
lm(TotalPrice ~ ColorD + ColorE + ColorF + ColorG + ColorH + ColorI,
   data = Diamonds)
```


> * **Interpretation:** A diamond with Color `D` compared to color `J` increases the expected total price by 3632.
> * **Interpretation:** A diamond with Color `E` compared to color `J` increases the expected total price by 2423



## Indicator variables {.small}

::: question
What is the reference category?
:::


```{r}
#| code-fold: false
lm(TotalPrice ~ ColorD + ColorE + ColorF + ColorG + ColorH + ColorI,
   data = Diamonds)
```


* **Interpretation:** A diamond with Color `D` compared to color `J` increases the expected total price by 3632.
* What is the interpretation for a diamond with Color `F`?



## R is smart

```{r}
#| code-fold: false
lm(TotalPrice ~ Color, data = Diamonds)
```


. . .

::: question
What is the reference category?
:::

## R is smart {.small}


```{r}
#| code-fold: false
lm(TotalPrice ~ Color, data = Diamonds)
```


> * What is the interpretation for Color `E` now?
> * What if we wanted a different referent category?
>   * We could code the indicators ourselves
>   * We could **relevel** the factor




## Relevel

```{r}
#| code-fold: false
levels(Diamonds$Color)
```

. . .

```{r}
#| code-fold: false
new_levels <- c("J", "D", "E", "F", "G", "H", "I")
Diamonds <- Diamonds %>%
  mutate(Color = fct_relevel(Color, new_levels))
```

```{r}
#| code-fold: false
levels(Diamonds$Color)
```



## R is smart


```{r}
#| code-fold: false
lm(TotalPrice ~ Color, data = Diamonds)
```


. . .

::: question
What is the reference category?
:::



## What if the variable is **binary**

* A **binary** variable is a special type of categorical variable with **two levels**



## ICU example

> * A sample of 200 patients in an ICU unit
> * Want to see if the patient's heart rate is related to whether they were admitted via the emergency room 
>   * y: Heart rate (beats per minute)
>   * x: indicator for emergency room admission
> * Aside: Is this inference or prediction?



## Binary x variable

```{r}
#| code-fold: false
data("ICU")
lm(Pulse ~ Emergency, data = ICU)
```

> * How can we interpret $\hat{\beta}_0$ now?
> * How can we interpret $\hat{\beta}_1$?



## {{< fa laptop >}} `Application Exercise` {.small}

1. Copy the following template into RStudio Pro:

```bash
https://github.com/sta-112-f22/appex-14.git
```

2. What are the variables in the `Diamonds` dataset?
3. What are the levels of the `Clarity` variable in the `Diamonds` data?
4. Fit a model with `TotalPrice` as the outcome and `Clarity` as the explanatory variable
5. Change the referent category to `SI1` and refit the model
6. Add the variable `Depth` to your model. How do you interpret the coefficient for this parameter?

```{r}
#| echo: false

countdown::countdown(5)
```


