---
title: "Assumptions of Logistic Regression"
author: "Lucy D'Agostino McGowan"
footer:  "[Dr. Lucy D'Agostino McGowan](https://lucymcgowan.com)"
logo: "img/favicon.png"
editor: source
format: 
  revealjs: 
    theme: [custom.scss]
    slide-number: true
    chalkboard: true
knitr:
  opts_chunk: 
    code-fold: true
    echo: true

---

## Conditions for simple linear regression {.small}

```{r child = "setup.Rmd"}
```
* Linearity
* Zero Mean
* Constant Variance
* Independence
* Random
* Normality




## Conditions for ~~simple~~ linear regression {.small}

* Linearity
* Zero Mean
* Constant Variance
* Independence
* Random
* Normality



## Conditions for linear regression {.small}

* **Linearity**
* Zero Mean
* Constant Variance
* Independence
* Random
* Normality


 
## Conditions for linear regression {.small}

::: question
How do we check these conditions?
::: 

* Linearity
* Zero Mean
* Constant Variance
* Independence
* Random
* Normality



## Conditions for linear regression {.small}

::: question
How do we check these conditions?
:::

Condition | Graph
----|----
Linearity | Residuals vs. fits
Zero Mean | _by default it's true_
Constant Variance | Residuals vs fits
Independence | No formal check
Random | No formal check
Normality | QQ-plot or histogram of residuals



## Conditions for **logistic** regression {.small}

* Linearity
* Independence
* Random



## Testing linearity for logistic regression {.small}

> * In logistic regression, the **log(odds)** are a linear function of $x$, that is: $\log(odds) \approx \beta_0 + \beta_1x$
> * You can check this by plotting the **empirical logits**
> * Example: `r emo::ji("golf")` Examining the relationship between the length of a putt with whether it was made

Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

::: question
What is the proportion of sucess when length is 3?
:::


Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

::: question
What is the proportion of sucesses when length is 3?
:::


Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134
Probability of success | 0.832 | 0.739 | 0.565 | 0.488 | 0.328



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

::: question
What are the odds of success when length is 3?
:::


Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134
Probability of success | 0.832 | 0.739 | 0.565 | 0.488 | 0.328



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

::: question
What are the odds of success when length is 3?
:::

Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134
Probability of success | 0.832 | 0.739 | 0.565 | 0.488 | 0.328
Odds of success | 4.941 | 2.839 | 1.298| 0.953| 0.489



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

::: question
What are the log(odds) of success when length is 3?
:::


Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134
Probability of success | 0.832 | 0.739 | 0.565 | 0.488 | 0.328
Odds of success | 4.941 | 2.839 | 1.298| 0.953| 0.489



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

::: question
What are the log(odds) of success when length is 3?
:::


Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134
Probability of success | 0.832 | 0.739 | 0.565 | 0.488 | 0.328
Odds of success | 4.941 | 2.839 | 1.298| 0.953| 0.489
Empirical logit | 1.60 | 1.04| 0.26 |−0.05 |−0.72



## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}

* We can plot the **empirical logit** to examine the linearity assumption

Length | 3 | 4 | 5 | 6 | 7
------|----|---|---|---|---
Number of successes | 84 | 88 | 61 | 61 | 44 
Number of failures | 17 | 31 | 47 | 64 | 90 
Total | 101 | 119 | 108 | 125 | 134
Probability of success | 0.832 | 0.739 | 0.565 | 0.488 | 0.328
Odds of success | 4.941 | 2.839 | 1.298| 0.953| 0.489
Empirical logit | 1.60 | 1.04| 0.26 |−0.05 |−0.72



## `r emo::ji("golf")` Testing for linearity in logistic regression  {.small}


```{r, fig.height = 2}
data <- data.frame(
  length = c(3, 4, 5, 6, 7),
  emp_logit = c(1.6, 1.04, 0.26, -0.05, -0.72)
)
ggplot(data, aes(length, emp_logit)) + 
  geom_point() + 
  labs(y = "log odds of success")
```




## `r emo::ji("golf")` Testing for linearity in logistic regression {.small}


```{r, fig.height = 2}
data <- data.frame(
  length = c(3, 4, 5, 6, 7),
  emp_logit = c(1.6, 1.04, 0.26, -0.05, -0.72)
)
ggplot(data, aes(length, emp_logit)) + 
  geom_point() + 
  geom_abline(intercept = 3.26, slope = -0.566, lty = 2) +
  labs(y = "log odds of success")
```




##  Testing for linearity in logistic regression {.small}

::: question
What if the $x$ variable isn't discrete?
:::

>* We can plot the **empirical logit** to examine the linearity assumption
>* You can create "bins" and calculate the empirical logit within each bin (for example, count the number of success when x is between 1 and 5: bin 1, count the number of successes when x is between 5 and 10: bin 2, etc)
