---
title: "Odds Ratios"
author: "Lucy D'Agostino McGowan"
footer:  "[Dr. Lucy D'Agostino McGowan](https://lucymcgowan.com)"
logo: "img/favicon.png"
editor: source
format: 
  revealjs: 
    theme: [custom.scss]
    slide-number: true
    chalkboard: true
knitr:
  opts_chunk: 
    code-fold: true
    echo: true
---


## Odds ratios {.small}

```{r child = "setup.Rmd"}
```



A study investigated whether a handheld device that sends a magnetic pulse into a person’s head
might be an effective treatment for migraine headaches.

* Researchers recruited 200 subjects who suffered from migraines 
* randomly assigned them to receive either the TMS (transcranial magnetic
stimulation) treatment or a placebo treatment 
* Subjects were instructed to apply the device at the onset of migraine symptoms and then assess how they felt two hours later. (either **Pain-free** or **Not pain-free**)



## Odds ratios {.small}

::: question
What is the explanatory variable?
:::

A study investigated whether a handheld device that sends a magnetic pulse into a person’s head might be an effective treatment for migraine headaches.

* Researchers recruited 200 subjects who suffered from migraines 
* randomly assigned them to receive either the TMS (transcranial magnetic
stimulation) treatment or a placebo treatment 
* Subjects were instructed to apply the device at the onset of migraine symptoms and then assess how they felt two hours later (either **Pain-free** or **Not pain-free**)


## Odds ratios {.small}

::: question
What type of variable is this?
:::

A study investigated whether a handheld device that sends a magnetic pulse into a person’s head might be an effective treatment for migraine headaches.

* Researchers recruited 200 subjects who suffered from migraines 
* randomly assigned them to receive either the TMS (transcranial magnetic
stimulation) treatment or a placebo treatment 
* Subjects were instructed to apply the device at the onset of migraine symptoms and then assess how they felt two hours later (either **Pain-free** or **Not pain-free**)


## Odds ratios {.small}

::: question
What is the outcome variable?
:::

A study investigated whether a handheld device that sends a magnetic pulse into a person’s head might be an effective treatment for migraine headaches.

* Researchers recruited 200 subjects who suffered from migraines 
* randomly assigned them to receive either the TMS (transcranial magnetic
stimulation) treatment or a placebo treatment 
* Subjects were instructed to apply the device at the onset of migraine symptoms and then assess how they felt two hours later (either **Pain-free** or **Not pain-free**)

 
## Odds ratios {.small}

::: question
What type of variable is this?
:::

A study investigated whether a handheld device that sends a magnetic pulse into a person’s head might be an effective treatment for migraine headaches.

* Researchers recruited 200 subjects who suffered from migraines 
* randomly assigned them to receive either the TMS (transcranial magnetic
stimulation) treatment or a placebo treatment 
* Subjects were instructed to apply the device at the onset of migraine symptoms and then assess how they felt two hours later (either **Pain-free** or **Not pain-free**)


## Odds ratios {.small}

|TMS | Placebo| Total
|-----|--------|------
Pain-free two hours later |39| 22 |61
Not pain-free two hours later |61 |78| 139
Total| 100| 100 |200

> * We can compare the results using **odds**
> * What are the odds of being pain-free for the placebo group?
>    * $(22/100)/(78/100) = 22/78 = 0.282$
> * What are the odds of being pain-free for the treatment group?
>    * $39/61 = 0.639$
> * Comparing the **odds** what can we conclude?
>    * TMS increases the likelihood of success


## Odds ratios {.small}

|TMS | Placebo| Total
|----|--------|------
Pain-free two hours later |39| 22 |61
Not pain-free two hours later |61 |78| 139
Total| 100| 100 |200

* We can summarize this relationship with an **odds ratio**: the ratio of the two odds

. . .

$OR = \frac{39/61}{22/78} = \frac{0.639}{0.282} = 2.27$

. . .

_"the odds of being pain free were 2.27 times higher with TMS than with the placebo"_



## Odds ratios {.small}

::: question
What if we wanted to calculate this in terms of _Not pain-free_ (with _pain-free_) as the referent?
:::

|TMS | Placebo| Total
|----|-------|-------
Pain-free two hours later |39| 22 |61
Not pain-free two hours later |61 |78| 139
Total| 100| 100 |200

. . .

$OR = \frac{61/39}{78/22} = \frac{1.564}{3.545} = 0.441$

. . .

_the odds for still being in pain for the TMS group are 0.441 times the odds of being in pain for the placebo group_
 


## Odds ratios {.small}

::: question
What changed here?
:::

|TMS | Placebo| Total
|-----|------|------
Pain-free two hours later |39| 22 |61
Not pain-free two hours later |61 |78| 139
Total| 100| 100 |200


$OR = \frac{78/22}{61/39} = \frac{3.545}{1.564} = 2.27$


_the odds for still being in pain for the placebo group are 2.27 times the odds of being in pain for the TMS group_
 



## Odds ratios {.small}

* In general, it's more natural to interpret odds ratios > 1, you can flip the referent to do so

|TMS | Placebo| Total
|-----|------|-------
Pain-free two hours later |39| 22 |61
Not pain-free two hours later |61 |78| 139
Total| 100| 100 |200


$OR = \frac{78/22}{61/39} = \frac{3.545}{1.564} = 2.27$


_the odds for still being in pain for the placebo group are 2.27 times the odds of being in pain for the TMS group_

<!-- insert exercise? -->

## Odds ratios {.small}

* Let's look at some Titanic data. We are interested in whether the sex of the passenger is related to whether they survived.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(Stat2Data)
data("Titanic")
options(digits = 3)
```

|Female | Male | Total
|-----|-------|------
Survived| 308 | 142 | 450
Died | 154 | 709| 863
Total | 462 | 851 | 1313



## Odds ratios {.small}

::: question
What are the odds of surviving for females versus males?
:::

* Let's look at some Titanic data. We are interested in whether the sex of the passenger is related to whether they survived.

|Female | Male | Total
|-------|-------|----
Survived| 308 | 142 | 450
Died | 154 | 709| 863
Total | 462 | 851 | 1313

. . .

$$OR = \frac{308/154}{142/709} = \frac{2}{0.2} = 9.99$$



## Odds ratios {.small}

::: question
How do you interpret this?
:::


|Female | Male | Total
|--------|-----|-------
Survived| 308 | 142 | 450
Died | 154 | 709| 863
Total | 462 | 851 | 1313


$$OR = \frac{308/154}{142/709} = \frac{2}{0.2} = 9.99$$


. . .

_the odds of surviving for the female passengers was 9.99 times the odds of surviving for the male passengers_



## Odds ratios {.small}

::: question
What if we wanted to fit a **model**? What would the equation be?
:::


|Female | Male | Total
|-|--|--
Survived| 308 | 142 | 450
Died | 154 | 709| 863
Total | 462 | 851 | 1313

. . . 


$$\log(\textrm{odds of survival}) = \beta_0 + \beta_1 \textrm{Male}$$



## Odds ratios {.small}


$$\log(\textrm{odds of survival}) = \beta_0 + \beta_1 \textrm{Male}$$


```{r}
#| code-fold: false
glm(Survived ~ Sex, data = Titanic, family = binomial)
```




## Odds Ratios {.small}

::: question
What is my referent category?
:::


$$\log(\textrm{odds of survival}) = \beta_0 + \beta_1 \textrm{Male}$$



```{r}
#| code-fold: false
levels(Titanic$Sex)
```



## Odds Ratios {.small}

::: question
How do I change that?
:::

$$\log(\textrm{odds of survival}) = \beta_0 + \beta_1 \textrm{Male}$$


```{r, eval = FALSE}
#| code-fold: false
Titanic <- Titanic %>%
  (Sex = ---(Sex, c("male", "female"))) 
```



## Odds Ratios {.small}

::: question
How do I change that?
:::

$$\log(\textrm{odds of survival}) = \beta_0 + \beta_1 \textrm{Male}$$

```{r}
#| code-fold: false
Titanic <- Titanic %>%
  mutate(Sex = fct_relevel(Sex, c("male", "female"))) 
```




## Odds Ratios {.small}

::: question
How do you interpret this result?
:::


```{r}
#| code-fold: false
model <- glm(Survived ~ Sex, data = Titanic, family = binomial)
summary(model)
```




## Odds Ratios {.small}

::: question
How do you interpret this result?
:::

::: small
```{r}
glm(Survived ~ Sex, data = Titanic, family = binomial) %>%
  summary()

exp(2.301176)
```
:::

_the odds of surviving for the female passengers was 9.99 times the odds of surviving for the male passengers_



## Odds ratios {.small}

> * What if the explanatory variable is continuous?
> * We did this already on Tuesday!

. . .

::: small
```{r}
data("MedGPA")
glm(Acceptance ~ GPA, data = MedGPA, family = binomial) %>%
  summary()
```
:::

. . .

_A one unit increase in GPA yields a 5.45 increase in the log odds of acceptance_




## Odds ratios {.small}

* What if the explanatory variable is continuous?
* We did this already on Tuesday!

::: small
```{r}
data("MedGPA")
glm(Acceptance ~ GPA, data = MedGPA, family = binomial) %>%
  summary()
```
:::

. . .

_A one unit increase in GPA yields a 234-fold increase in the odds of acceptance_

. . .

* `r emo::ji("scream")` that seems huge! **Remember:** the interpretation of these coefficients depends on your units (the same as in ordinary linear regression).



## Odds ratios {.small}

::: question
How could we get the odds associated with increasing GPA by 0.1?
:::

::: small
```{r}
glm(Acceptance ~ GPA, data = MedGPA, family = binomial) %>%
  summary()
```
:::

. . .

```{r}
exp(5.454) ## a one unit increase in GPA
exp(5.454 * 0.1) ## a 0.1 increase in GPA
```


. . .

_A one-tenth unit increase in GPA yields a 1.73-fold increase in the odds of acceptance_



## Odds ratios {.small}

::: question
How could we get the odds associated with increasing GPA by 0.1?
:::

::: small
```{r}
#| code-fold: false
MedGPA <- MedGPA %>%
  mutate(GPA_10 = GPA * 10)

glm(Acceptance ~ GPA_10, data = MedGPA, family = binomial) %>%
  summary()
```
:::

_A one-tenth unit increase in GPA yields a 1.73-fold increase in the odds of acceptance_
